---
title: "Supplementary Material for 'Estimating the effect of a scanner upgrade on measures of grey matter structure for longitudinal designs'"
author:
  - Evelyn Medawar, MSc [1][2][3]
  - Ronja Thieleking, MSc [1]
  - Iryna Manuilova, BSc [1]
  - Maria Paerisch[1]
  - Arno Villringer, Prof. [1][2][3][4][5]
  - A. Veronica Witte, PhD [1][4][5]
  - Frauke Beyer, PhD [1][4]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline 2 Berlin School of Mind and Brain, Humboldt-Universität zu Berlin, Berlin, Germany \newline 3 Center for Stroke Research Berlin (CSB), Charité Universitätsmedizin, Berlin, Germany \newline 4 CRC 1052 “Obesity Mechanisms”, Subproject A1, Leizpig University, Leipzig, Germany \newline 5 Day Clinic for Cognitive Neurology, University of Leipzig Medical Center - Leipzig University, Leipzig, Germany

output:
  bookdown::pdf_book: 
    fig_caption: yes
    always_allow_html: true
    number_sections: yes
    toc: yes
    extra_dependencies: ["float"]
bibliography: "bibliography.bib" 

---

```{r "Package loading", include=FALSE}
library(lme4)
library(ggplot2)
library(haven)
library(plyr)
library(dplyr)
library(tidyr)
library(knitr)
library(psych)
library(psy)
library(boot)
library(MuMIn)
library(kableExtra)
library(patchwork)
library(wesanderson)
library(lmerTest)
library(lsr)
options(knitr.table.format = "latex")
library(cowplot)
source("icc_table.R")
source("create_long_table.R")
source("cnr_rh_table.R")
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.height=2, fig.width=4)


library(fsbrain)
atlas = 'aparc';
template_subject = 'fsaverage';
template_subjects_dir = "/afs/cbs.mpg.de/software/freesurfer/6.0.0p1/ubuntu-bionic-amd64/subjects/"; #your FS directory 
library('rgl');
r3dDefaults$windowRect=c(50, 50, 500, 500);

#makecmap_th = colorRampPalette(RColorBrewer::brewer.pal(9, name="YlOrRd"));
makecmap_pd = colorRampPalette(RColorBrewer::brewer.pal(9, name="RdBu"));



aparc_ct=read.csv("data/aparc_ct_update.csv")
aparc_area=read.csv("data/aparc_area_update.csv")
aparc_volume=read.csv("data/aparc_volume_update.csv")
```

```{r create thickness table, echo=FALSE}
aparc_ct_l=create_long_table_aparc(aparc_ct,"thickness")
#---- calculate ICC and PD for FreeSurfer cortical thickness----
icc_res_th<-create_icc_table_aparc(aparc_ct_l)
colnames(icc_res_th)=c("ROI","hemi","ICC", "lower ICC", "upper ICC", "PD", "T", "p", "adj.p")
```

```{r create area table, echo=FALSE}
aparc_area_l=create_long_table_aparc(aparc_area, "area")
#---- calculate ICC and PD for FreeSurfer cortical area----
icc_res_area<-create_icc_table_aparc(aparc_area_l)
colnames(icc_res_area)=c("ROI","hemi","ICC", "lower ICC", "upper ICC", "PD", "T", "p", "adj.p")
```

```{r create volume table, echo=FALSE}
aparc_volume_l=create_long_table_aparc(aparc_volume,"volume")
#---- calculate ICC and PD for FreeSurfer cortical volume----
icc_res_vol<-create_icc_table_aparc(aparc_volume_l)
colnames(icc_res_vol)=c("ROI","hemi","ICC", "lower ICC", "upper ICC", "PD", "T", "p", "adj.p")
```

```{r create QA measures table, echo=FALSE}
qa=read.csv("data/mriqc_measures.csv")
for (i in 1:nrow(qa)){
  #print(i)
  #print(qa[i,1])
  tmp=strsplit(toString(qa[i,1]),'_')[[1]]
  #print(tmp)
  qa[i,"subj"]=tmp[1]
  qa[i,"scanner"]=tmp[2]
  qa[i,"acq"]=tmp[3]
}

qa$Scanner=as.factor(qa$scanner)
levels(qa$Scanner)=c("Skyra","Verio")
qa$Acquisition=as.factor(qa$acq)
levels(qa$Acquisition)=c("D","ND")
```

```{r correlation of PD and CNR, echo=FALSE}
for (i in 1:nrow(aparc_ct_l[1:418,])){
  aparc_ct_l[i,"subj"]=paste("sub-",strsplit(toString(aparc_ct_l[i,"subj.ID"]),'[.]')[[1]][1],strsplit(toString(aparc_ct_l[i,"subj.ID"]),'[.]')[[1]][2],sep="")
}
aparc_ct_l[419:464,"subj"]=paste("sub-",aparc_ct_l[419:464,"subj.ID"],sep="")
aparc_ct_l$scanner_m=aparc_ct_l$scanner

qa[qa$scanner=="ses-VER1","scanner_m"]="VERIO"
qa[qa$scanner=="ses-SKYRA"&qa$acq=="acq-D","scanner_m"]="SKYRA"
qa$subj.ID=qa$subj
qa$scanner_m=qa$scanner
merged_qa=merge(qa[!is.na(qa$scanner_m),c("subj.ID","scanner_m","cnr")],aparc_ct_l, by=c("subj.ID","scanner_m"))
cnr_th_table=create_cnr_th_table(merged_qa)
```

```{r create unwarped thickness table, echo=FALSE}
aparc_unw=read.csv("data/aparc_ct_unw.csv")
aparc_ct_l=create_long_table_aparc(aparc_unw, "thickness")

#---- calculate ICC and PD for FreeSurfer cortical thickness----
icc_res_th_unw<-create_icc_table_aparc(aparc_ct_l)
colnames(icc_res_th_unw)=c("ROI","hemi","ICC", "lower ICC", "upper ICC", "PD", "T", "p", "adj.p")
```



\newpage


# Bland-Altman plot for all subcortical ROI
```{r blandaltmanallinone, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Bland-Altman plot showing differences of Verio-Skyra against means for all subcortical ROI. Left/right panel show left/right hemisphere.)",fig.show='hold',fig.align='center'}
Subcort_BA_allinone=("figures/Bland_Altmann_plot_subcort_allinone.png")
knitr::include_graphics(Subcort_BA_allinone)
```
\newpage


# Cortical thickness 
## Vertex-wise analysis of cortical thickness ICC
```{r vertexthicc, echo=FALSE,out.width="49%", out.height="30%",fig.cap="ICC for cortical thickness (CT) on white surface (left panel) and inflated surface (right panel)",fig.show='hold',fig.align='center'}
thickness_ICC=c("figures/pial/thickness_sm10_ICC.png","figures/thickness_sm10_ICC.png")
knitr::include_graphics(thickness_ICC)
```
## Vertex-wise analysis of cortical thickness PD

```{r vertexthpd, echo=FALSE,out.width="49%", out.height="30%",fig.cap="PD for cortical thickness (CT) on white surface (left panel) and inflated surface (right panel). Positive values: Verio>Skyra , negative values: Verio<Skyra ))",fig.show='hold',fig.align='center'}
pd_th=c("figures/pial/thickness_sm10_pd.png","figures/thickness_sm10_pd.png")
knitr::include_graphics(pd_th)
```
## Reliability and percent difference table of cortical thickness

```{r tab1thickness, echo=FALSE}
table_th=
  icc_res_th %>%
    mutate(
      adj.p = cell_spec(adj.p, "latex", bold = ifelse(adj.p < 0.05, T, F))
    )  %>%
   kable("latex", escape = F, caption="Reliability and differences for cortical thickness") %>%
   kable_styling(bootstrap_options = c("striped", "hover"))
table_th
```
\newpage
# Cortical area 
## Vertex-wise analysis of cortical area ICC

```{r vertexareaicc, echo=FALSE,out.width="49%", out.height="30%",fig.cap="ICC for cortical area (CA) on white surface (left panel) and inflated surface (right panel)",fig.show='hold',fig.align='center'}
area_ICC=c("figures/pial/area_sm10_ICC.png","figures/area_sm10_ICC.png")
knitr::include_graphics(area_ICC)
```
## Vertex-wise analysis of cortical area PD

```{r vertexareapd, echo=FALSE,out.width="49%", out.height="30%",fig.cap="PD for cortical area (CA) on white surface (left panel) and inflated surface (right panel). Positive values: Verio>Skyra , negative values: Verio<Skyra ))",fig.show='hold',fig.align='center'}
pd_area=c("figures/pial/area_sm10_pd.png","figures/area_sm10_pd.png")
knitr::include_graphics(pd_area)
```
## Reliability and percent difference table of cortical area

```{r tab2area,echo=FALSE}
table_area=
icc_res_area %>%
   mutate(
     adj.p = cell_spec(adj.p, "latex", bold = ifelse(adj.p < 0.05, T, F))
   )  %>%
  #select(p_adjust) %>%
  kable("latex", escape = F, caption="Reliability and differences for cortical area") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
table_area
```
\newpage
# Cortical volume 
## Vertex-wise analysis of cortical volume ICC

```{r vertexvolumeicc, echo=FALSE,out.width="49%", out.height="30%",fig.cap="ICC for cortical volume (CV) on white surface (left panel) and inflated surface (right panel)",fig.show='hold',fig.align='center'}
volume_ICC=c("figures/pial/volume_sm10_ICC.png","figures/volume_sm10_ICC.png")
knitr::include_graphics(volume_ICC)
```
## Vertex-wise analysis of cortical volume PD

```{r vertexvolumepd, echo=FALSE, out.width="49%", out.height="30%",fig.cap="PD for cortical volume (CV) on white surface (left panel) and inflated surface (right panel). Positive values: Verio>Skyra , negative values: Verio<Skyra ))",fig.show='hold',fig.align='center'}
pd_volume=c("figures/pial/volume_sm10_pd.png","figures/volume_sm10_pd.png")
knitr::include_graphics(pd_volume)
```
## Reliability and percent difference table of cortical volume

```{r tab3volume,echo=FALSE}
table_vol=
  icc_res_vol %>%
   mutate(
     adj.p = cell_spec(adj.p, "latex", bold = ifelse(adj.p < 0.05, T, F))
   )  %>%
  #select(p_adjust) %>%
  kable("latex", escape = F, caption="Reliability and differences for cortical volume") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
table_vol
```
\newpage
# Correlation of CNR and CT

```{r cnrcttab, echo=FALSE}
colnames(cnr_th_table)=c("ROI","hemi","CNR estimate", "scanner estimate", "p CNR","p scanner", "adj.p.CNR","adj.p.scanner")
cnr_th_table[,c(-1,-2)]=round(cnr_th_table[,c(-1,-2)],3)
cnrcttab=
  cnr_th_table %>%
   mutate(
     adj.p.CNR = cell_spec(adj.p.CNR, "latex", bold = ifelse(adj.p.CNR < 0.05, T, F))
   )  %>%
  mutate(
     adj.p.scanner = cell_spec(adj.p.scanner, "latex", bold = ifelse(adj.p.scanner < 0.05, T, F))
   )   %>%
  kable("latex", escape = F, caption="Association of CNR and scanner with CT") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
cnrcttab
```
\newpage

# Gradunwarp gradient distortion corrected data
## Vertex-wise analysis of gradunwarp distortion-corrected ICC
```{r vertexthiccunwarped, echo=FALSE,out.width="49%", out.height="30%",fig.cap="ICC for cortical thickness (CT) of gradunwarp Skyra ND and Verio ND on white surface (left panel) and inflated surface (right panel)",fig.show='hold',fig.align='center'}
thickness_ICC=c("figures/unwarped/thickness_sm10_ICC.png", "figures/unwarped/thickness_sm10_ICC_inf.png")
knitr::include_graphics(thickness_ICC)
```
## Vertex-wise analysis of gradunwarp distortion-corrected PD

```{r vertexthpdunwarped, echo=FALSE,out.width="49%", out.height="30%",fig.cap="PD for cortical thickness (CT) of gradunwarp Skyra ND and Verio ND on white surface (left panel) and inflated surface (right panel) Positive values: Verio>Skyra , negative values: Verio<Skyra ))",fig.show='hold',fig.align='center'}
pd_th=c("figures/unwarped/thickness_sm10_pd.png","figures/unwarped/thickness_sm10_pd_inf.png")
knitr::include_graphics(pd_th)
```

```{r tab5thunwarped, echo=FALSE}
table_th_unw=
icc_res_th_unw %>%
    mutate(
      adj.p = cell_spec(adj.p, "latex", bold = ifelse(adj.p < 0.05, T, F))
    )  %>%
   kable("latex", escape = F, caption="Reliability and percent difference for CT from gradunwarp distortion corrected data (T<0 reflects Skyra >Verio , T>0 reflects Verio >Skyra )") %>%
   kable_styling(bootstrap_options = c("striped", "hover"))
table_th_unw
```

# Skyra D versus Verio ND analysis
```{r create former main (skyra_d vs verio_nd) thickness table, echo=FALSE}
#there is none complete data for some subjects, which are probably running right now.
aparc_veriond_skyrad=read.csv("data/aparc_ct_veriond_skyrad.csv")
aparc_ct_l=create_long_table_aparc(aparc_veriond_skyrad, "thickness")

#---- calculate ICC and PD for FreeSurfer cortical thickness----
icc_res_th_main<-create_icc_table_aparc(aparc_ct_l)
colnames(icc_res_th_main)=c("ROI","hemi","ICC", "lower ICC", "upper ICC", "PD", "T", "p", "adj.p")
```

```{r tab6main, echo=FALSE}
table_th_main=
icc_res_th_main %>%
    mutate(
      adj.p = cell_spec(adj.p, "latex", bold = ifelse(adj.p < 0.05, T, F))
    )  %>%
   kable("latex", escape = F, caption="Reliability and percent difference for cortical thickness from Skyra D versus Verio ND data (T<0 reflects Skyra >Verio , T>0 reflects Verio >Skyra )") %>%
   kable_styling(bootstrap_options = c("striped", "hover"))
table_th_main
```

# QA differences

```{r plotcnruncorrwskyra,fig.cap="CNR differences between the Skyra D and ND acquisitions (left panel) and between Skyra D and Verio ND acquisitions (right panel), showing higher CNR irrespective of gradient distortion on the Verio scanner", fig.width=6, fig.height=4,  echo=FALSE}
tspag_wSkyra = ggplot(qa[(qa$scanner=="ses-SKYRA"),], aes(x=Acquisition, y=cnr, fill=Acquisition)) + 
  geom_boxplot()+
  labs(subtitle="CNR differences between \nSkyra D and ND acquisitions")+
  xlab(bquote("Acquisition")) + 
  ylab(bquote('CNR')) + 
  scale_fill_manual(values=wes_palette("Moonrise2",2,type="discrete")) +
  theme(axis.text=element_text(size=10),
  axis.title=element_text(size=12),
  strip.text = element_text(size=12),
  legend.position="top")

tspag_ND = ggplot(qa[!(qa$Acquisition=="ND"),], aes(x=Scanner, y=cnr, fill=Scanner)) + 
  geom_boxplot()+
  labs(subtitle="CNR differences between Skyra D\nand Verio ND acquisitions")+
  scale_fill_manual(values=wes_palette("Moonrise2",2,type="discrete")) +
  xlab(bquote("Scanner")) + 
  ylab(bquote('CNR')) + 
  theme(axis.text=element_text(size=10),
  axis.title=element_text(size=12),
  strip.text = element_text(size=12),
  legend.position="top")

pl2=(tspag_wSkyra|tspag_ND)
pl2

ggsave("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/Analysis_R/Figures/PLOS/Revision/SuppFig1.eps", pl2, dpi=300,width=14, units = "cm")
```

```{r difference between acquisitions on same scanner, echo=FALSE}
res_cnr=lmer(cnr ~ acq + (1|subj), data=qa[(qa$scanner=="SKYRA"),]) 
null_cnr=lmer(cnr ~ (1|subj), data=qa[(qa$scanner=="SKYRA"),]) 
anova_cnr_Skyra =anova(res_cnr,null_cnr)

p_Skyra_cnr=anova_cnr_Skyra$`Pr(>Chisq)`[2]
```
When comparing the acquisitions with and without vendor-provided online gradient distortion correction on the Skyra scanner (D and ND), we observed that the distortion correction increased CNR ($\beta$=`r round(fixef(res_cnr)[2],3)`, p < 0.001, see Figure \@ref(fig:plotcnruncorrwskyra), left panel). 

```{r difference between ND acquisitions verio/skyra, echo=FALSE}
res_cnr=lmer(cnr ~ scanner + (1|subj), data=qa[(qa$acq=="D"),]) 
null_cnr=lmer(cnr ~ (1|subj), data=qa[(qa$acq=="D"),]) 
anov_ND_cnr=anova(res_cnr,null_cnr)
p_ND_cnr=anov_ND_cnr$`Pr(>Chisq)`[2]

res_cjv=lmer(cjv ~ scanner + (1|subj), data=qa[(qa$acq=="D"),]) 
null_cjv=lmer(cjv ~ (1|subj), data=qa[(qa$acq=="D"),]) 
anov_ND_cjv=anova(res_cjv,null_cjv)
p_ND_cjv=anov_ND_cjv$`Pr(>Chisq)`[2]

res_efc=lmer(efc ~ scanner + (1|subj), data=qa[(qa$acq=="D"),]) 
null_efc=lmer(efc ~ (1|subj), data=qa[(qa$acq=="D"),]) 
anov_ND_efc=anova(res_efc,null_efc)
p_ND_efc=anov_ND_efc$`Pr(>Chisq)`[2]
```
When comparing Verio ND and Skyra D, we also see higher CNR on the Verio scanner ($\beta$=`r round(fixef(res_cnr)[2],3)`, p < 0.001, see Figure \@ref(fig:plotcnruncorrwskyra), right panel), which is expected given the CNR difference between Verio ND and Skyra ND reported in the main manuscript.