---
title: "Supplementary Material for 'Estimating the effect of a scanner upgrade on measures of gray matter structure for longitudinal designs'"
author:
  - Evelyn Medawar, MSc, [1]
  - Ronja Thieleking, MSc, [1]
  - Iryna Manuilova, BSc, [1]
  - Maria Paerisch,[1]
  - Arno Villringer, Prof., [1][2][3]
  - A. Veronica Witte, PhD, [1][2]
  - Frauke Beyer, PhD, [1][2]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline 2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3 Day Clinic for Cognitive Neurology, University Clinic Leipzig
output:
  bookdown::pdf_book: 
    fig_caption: yes
    always_allow_html: true
    number_sections: yes
    toc: yes
bibliography: "bibliography.bib" 

---

```{r "Package loading", include=FALSE}
library(lme4)
library(ggplot2)
library(haven)
library(plyr)
library(dplyr)
library(tidyr)
library(knitr)
library(psych)
library(psy)
library(boot)
library(MuMIn)
library(kableExtra)
library(patchwork)
library(lsr)
options(knitr.table.format = "latex")
library(cowplot)
source("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/Analysis_R/icc_table.R")
source("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/Analysis_R/create_long_table.R")
source("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/Analysis_R/cnr_rh_table.R")
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, fig.height=2, fig.width=4)


library(fsbrain)
atlas = 'aparc';
template_subject = 'fsaverage';
template_subjects_dir = "/data/pt_nro186_lifeupdate/Data/FREESURFER/FREESURFER_long/"; 
library('rgl');
r3dDefaults$windowRect=c(50, 50, 500, 500);

#makecmap_th = colorRampPalette(RColorBrewer::brewer.pal(9, name="YlOrRd"));
makecmap_pd = colorRampPalette(RColorBrewer::brewer.pal(9, name="RdBu"));
```

```{r create thickness table, echo=FALSE}
aparc_ct_rh=read.csv("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/FS_stats_allsubjects/longitudinal_statistics/cort/DK_thickness_VERIO_ND_SKYRA_D_rh_main.txt", sep = '\t')
aparc_ct_lh=read.csv("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/FS_stats_allsubjects/longitudinal_statistics/cort/DK_thickness_VERIO_ND_SKYRA_D_lh_main.txt", sep = '\t')
aparc_ct=merge(aparc_ct_rh,aparc_ct_lh, by.x="rh.aparc.thickness", by.y="lh.aparc.thickness")
data=merge(aparc_ct_rh,aparc_ct_lh, by.x="rh.aparc.thickness", by.y="lh.aparc.thickness")
colnames(aparc_ct)[1]="ID"
aparc_ct_l=create_long_table_aparc(aparc_ct,"thickness")
#---- calculate ICC and PD for FreeSurfer cortical thickness----
icc_res_th<-create_icc_table_aparc(aparc_ct_l)
colnames(icc_res_th)=c("ROI","hemi","ICC", "lower ICC", "upper ICC", "PD", "T", "p", "adj.p")
```

```{r create area table, echo=FALSE}
aparc_area_rh=read.csv("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/FS_stats_allsubjects/longitudinal_statistics/cort/DK_area_VERIO_ND_SKYRA_D_rh_main.txt", sep = '\t')
aparc_area_lh=read.csv("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/FS_stats_allsubjects/longitudinal_statistics/cort/DK_area_VERIO_ND_SKYRA_D_lh_main.txt", sep = '\t')
aparc_area=merge(aparc_area_rh,aparc_area_lh, by.x="rh.aparc.area", by.y="lh.aparc.area")
colnames(aparc_area)[1]="ID"
aparc_area_l=create_long_table_aparc(aparc_area, "area")
#---- calculate ICC and PD for FreeSurfer cortical area----
icc_res_area<-create_icc_table_aparc(aparc_area_l)
colnames(icc_res_area)=c("ROI","hemi","ICC", "lower ICC", "upper ICC", "PD", "T", "p", "adj.p")
```

```{r create volume table, echo=FALSE}
aparc_volume_rh=read.csv("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/FS_stats_allsubjects/longitudinal_statistics/cort/DK_volume_VERIO_ND_SKYRA_D_rh_main.txt", sep = '\t')
aparc_volume_lh=read.csv("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/FS_stats_allsubjects/longitudinal_statistics/cort/DK_volume_VERIO_ND_SKYRA_D_lh_main.txt", sep = '\t')
aparc_volume=merge(aparc_volume_rh,aparc_volume_lh, by.x="rh.aparc.volume", by.y="lh.aparc.volume")
colnames(aparc_volume)[1]="ID"
aparc_volume_l=create_long_table_aparc(aparc_volume,"volume")
#---- calculate ICC and PD for FreeSurfer cortical volume----
icc_res_vol<-create_icc_table_aparc(aparc_volume_l)
colnames(icc_res_vol)=c("ROI","hemi","ICC", "lower ICC", "upper ICC", "PD", "T", "p", "adj.p")
```

```{r create QA measures table, echo=FALSE}
qa=read.table('/data/pt_nro186_lifeupdate/Data/BIDS/derivatives/mriqc/group_T1w.tsv', header=T)
for (i in 1:nrow(qa)){
  #print(i)
  #print(qa[i,1])
  tmp=strsplit(toString(qa[i,1]),'_')[[1]]
  #print(tmp)
  qa[i,"subj"]=tmp[1]
  qa[i,"scanner"]=tmp[2]
  qa[i,"acq"]=tmp[3]
}

```

```{r correlation of PD and CNR, echo=FALSE}
for (i in 1:nrow(aparc_ct_l[1:418,])){
  aparc_ct_l[i,"subj"]=paste("sub-",strsplit(toString(aparc_ct_l[i,"subj.ID"]),'[.]')[[1]][1],strsplit(toString(aparc_ct_l[i,"subj.ID"]),'[.]')[[1]][2],sep="")
}
aparc_ct_l[419:464,"subj"]=paste("sub-",aparc_ct_l[419:464,"subj.ID"],sep="")
aparc_ct_l$scanner_m=aparc_ct_l$scanner

qa[qa$scanner=="ses-VER1","scanner_m"]="VERIO"
qa[qa$scanner=="ses-SKYRA"&qa$acq=="acq-D","scanner_m"]="SKYRA"

merged_qa=merge(qa[!is.na(qa$scanner_m),c("subj","scanner_m","cnr")],aparc_ct_l, by=c("subj","scanner_m"))
cnr_th_table=create_cnr_th_table(merged_qa)
```

```{r create unwarped thickness table, echo=FALSE}
aparc_ct_rh=read.csv("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/FS_stats_allsubjects/longitudinal_statistics/cort/DK_thickness_VERIO_ND_SKYRA_D_rh_unwarped.txt", sep = '\t')
aparc_ct_lh=read.csv("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/FS_stats_allsubjects/longitudinal_statistics/cort/DK_thickness_VERIO_ND_SKYRA_D_lh_unwarped.txt", sep = '\t')
aparc_ct=merge(aparc_ct_rh,aparc_ct_lh, by.x="rh.aparc.thickness", by.y="lh.aparc.thickness", all.x)
colnames(aparc_ct)[1]="ID"
#there is none complete data for some subjects, which are probably running right now.
aparc_ct_l=create_long_table_aparc(aparc_ct, "thickness")

#---- calculate ICC and PD for FreeSurfer cortical thickness----
icc_res_th_unw<-create_icc_table_aparc(aparc_ct_l)
colnames(icc_res_th_unw)=c("ROI","hemi","ICC", "lower ICC", "upper ICC", "PD", "T", "p", "adj.p")
```

```{r create subcort table, echo=FALSE}
vol=read.table('/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/FS_stats_allsubjects/longitudinal_statistics/subcort/aseg_stats.txt', header=T)
vol_l=create_long_table_subcort(vol)
icc_res_vol<-create_icc_table_vol(vol_l)
colnames(icc_res_vol)=c("ROI","hemi","ICC", "lower ICC", "upper ICC", "PD", "T", "p", "adj.p")
```

# Cortical thickness 
## CT reliability and percent difference table

```{r tab1thickness, echo=FALSE}
table_th=
  icc_res_th %>%
    mutate(
      adj.p = cell_spec(adj.p, "latex", bold = ifelse(adj.p < 0.05, T, F))
    )  %>%
   kable("latex", escape = F, caption="Reliability and differences for cortical thickness") %>%
   kable_styling(bootstrap_options = c("striped", "hover"))
table_th
```

## Vertex-wise analysis of cortical thickness ICC
```{r vertexthicc, echo=FALSE,out.width="49%", out.height="30%",fig.cap="Lateral and medial view of ICC for thickness (left, right panel)",fig.show='hold',fig.align='center'}
thickness_ICC=c("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/ICC_analysis/ICC_pairedt_results/lh.thickness_sm10_ICC_lateral.png","/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/ICC_analysis/ICC_pairedt_results/lh.thickness_sm10_ICC_medial.png")
knitr::include_graphics(thickness_ICC)
```
## Vertex-wise analysis of cortical thickness PD
```{r vertexthpd, echo=FALSE,out.width="49%", out.height="30%",fig.cap="Lateral and medial view of PD for thickness (positive values: Verio >Skyra , negative values: Verio <Skyra ))",fig.show='hold',fig.align='center'}
pd_th=c("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/ICC_analysis/ICC_pairedt_results/lh.thickness_sm10_pd_lateral.png","/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/ICC_analysis/ICC_pairedt_results/lh.thickness_sm10_pd_medial.png")
knitr::include_graphics(pd_th)
```

# Cortical area 
## Reliability and percent difference table
```{r tab2area,echo=FALSE}
table_area=
icc_res_area %>%
   mutate(
     adj.p = cell_spec(adj.p, "latex", bold = ifelse(adj.p < 0.05, T, F))
   )  %>%
  #select(p_adjust) %>%
  kable("latex", escape = F, caption="Reliability and differences for cortical area") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
table_area
```

## Vertex-wise analysis of cortical area ICC
```{r vertexareaicc, echo=FALSE,out.width="49%", out.height="30%",fig.cap="Lateral and medial view of ICC for area (left, right panel)",fig.show='hold',fig.align='center'}
area_ICC=c("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/ICC_analysis/ICC_pairedt_results/lh.area_sm10_ICC_lateral.png","/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/ICC_analysis/ICC_pairedt_results/lh.area_sm10_ICC_medial.png")
knitr::include_graphics(area_ICC)
```
## Vertex-wise analysis of cortical area PD
```{r vertexareapd, echo=FALSE,out.width="49%", out.height="30%",fig.cap="Lateral and medial view of PD for area (positive values: Verio >Skyra , negative values: Verio <Skyra )",fig.show='hold',fig.align='center'}
pd_area=c("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/ICC_analysis/ICC_pairedt_results/lh.area_sm10_pd_lateral.png","/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/ICC_analysis/ICC_pairedt_results/lh.area_sm10_pd_medial.png")
knitr::include_graphics(pd_area)
```

# Cortical volume 
## Reliability and percent difference table
```{r tab3volume,echo=FALSE}
table_vol=
  icc_res_vol %>%
   mutate(
     adj.p = cell_spec(adj.p, "latex", bold = ifelse(adj.p < 0.05, T, F))
   )  %>%
  #select(p_adjust) %>%
  kable("latex", escape = F, caption="Reliability and differences for cortical volume") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
table_vol
```
## Vertex-wise analysis of cortical volume ICC
```{r vertexvolumeicc, echo=FALSE,out.width="49%", out.height="30%",fig.cap="Lateral and medial view of ICC for volume (left, right panel)",fig.show='hold',fig.align='center'}
volume_ICC=c("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/ICC_analysis/ICC_pairedt_results/lh.volume_sm10_ICC_lateral.png","/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/ICC_analysis/ICC_pairedt_results/lh.volume_sm10_ICC_medial.png")
knitr::include_graphics(volume_ICC)
```
## Vertex-wise analysis of cortical volume PD
```{r vertexvolumepd, echo=FALSE,out.width="49%", out.height="30%",fig.cap="Lateral and medial view of PD for volume (positive values: Verio >Skyra , negative values: Verio <Skyra )",fig.show='hold',fig.align='center'}
pd_volume=c("/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/ICC_analysis/ICC_pairedt_results/lh.volume_sm10_pd_lateral.png","/data/pt_nro186_lifeupdate/Results/GMV_project_evelyn_frauke/FS_Group_Analysis/ICC_analysis/ICC_pairedt_results/lh.volume_sm10_pd_medial.png")
knitr::include_graphics(pd_volume)
```

# Correlation of CNR and CT
```{r cnrcttab, echo=FALSE}
colnames(cnr_th_table)=c("ROI","hemi","CNR estimate", "scanner estimate", "p CNR","p scanner", "adj.p.CNR","adj.p.scanner")
cnr_th_table[,c(-1,-2)]=round(cnr_th_table[,c(-1,-2)],3)
cnrcttab=
  cnr_th_table %>%
   mutate(
     adj.p.CNR = cell_spec(adj.p.CNR, "latex", bold = ifelse(adj.p.CNR < 0.05, T, F))
   )  %>%
  mutate(
     adj.p.scanner = cell_spec(adj.p.scanner, "latex", bold = ifelse(adj.p.scanner < 0.05, T, F))
   )   %>%
  kable("latex", escape = F, caption="Association of CNR and scanner with CT") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
cnrcttab
```

# Cortical thickness reliability and percent difference table for gradunwarp distortion corrected data
```{r tab5thunwarped, echo=FALSE}
table_th_unw=
icc_res_th_unw %>%
    mutate(
      adj.p = cell_spec(adj.p, "latex", bold = ifelse(adj.p < 0.05, T, F))
    )  %>%
   kable("latex", escape = F, caption="Reliability and percent difference for cortical thickness from gradient non-linearity corrected data (T<0 reflects Skyra >Verio , T>0 reflects Verio >Skyra )") %>%
   kable_styling(bootstrap_options = c("striped", "hover"))
table_th_unw
```
